---
title: "DSAIT4200 Coursework B --- A Transfer Learning Experiment"
author: "Melani Evangelou, Luca Ionescu, Chris Obis, Maksym Ziemlewski"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
  pdf_document:
    toc: true
subtitle: Reproducible analysis
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 3, dpi = 130)
```

# Purpose and approach

The experiment measures a **metric score in `[0, 1]`** for nine models
on seven test datasets (`TeD`). Transfer-learning models are trained on
subsets of eight training datasets (`TrD1..TrD8`). We answer:

- **RQ1** - does transfer learning beat non-transfer baselines?
- **RQ2** - how do the **number** and **identity** of training datasets,
  and the **depth of refinement**, affect performance?

To achieve this, we will: (1) explore the data, (2) fit a **large
preliminary model and rank effects with Type-II ANOVA** to guide the
analysis, (3) **transform the response per test dataset** (within-`TeD`
standardization) so effects are comparable across test sets, (4) build
structure-aware models and report **effect sizes with confidence
intervals** (in SD units), and (5) confirm everything with a
bounded-response **logit / quasi-binomial** model.

```{r libs}
# install.packages(c("tidyverse","car","emmeans","broom","patchwork"))
library(patchwork); library(tidyverse); library(car); library(emmeans); library(broom)
theme_set(theme_bw(base_size = 11)); emm_options(rg.limit = 200000)
```

# Data preparation

We derive the non-explicit factors: the **number** of training datasets
`n_trd`, the architecture `model_type`, `model_group` (Baseline vs
Transfer), the broad `task_type`, the **refinement depth** of the
multi-`TrD` models (`MN`=0 ... `MF`=4), the **within-`TeD`
standardized** score `z` (primary response, SD-of-that-TeD units), and a
clamped **logit** for the bounded-response robustness check.

```{r prep}
d <- readr::read_csv("data.csv", show_col_types = FALSE)
trd_cols <- paste0("TrD", 1:8); eps <- 0.005

d <- d |>
  mutate(
    n_trd       = rowSums(across(all_of(trd_cols))),
    model_group = if_else(model %in% c("B1","B2","B3"), "BL", "TR"),
    model_type  = if_else(model %in% c("B1","B2","B3"), "Baseline", model),
    task_type   = case_when(TeD %in% paste0("TeD",1:4) ~ "Classification",
                            TeD == "TeD5" ~ "Recommendation", TRUE ~ "Regression"),
    depth       = case_when(model_type=="MN"~0, model_type=="M1"~1, model_type=="M2"~2,
                            model_type=="M3"~3, model_type=="MF"~4, TRUE~NA_real_),
    logit       = qlogis(pmin(pmax(score, eps), 1 - eps))) |> # mapping (0,1) to the whole number line
  
  group_by(TeD) |> mutate(z = (score - mean(score)) / sd(score)) |> ungroup() |>
  mutate(TeD        = factor(TeD, levels = paste0("TeD",1:7)),
         model_group= factor(model_group, levels = c("BL","TR")),
         model_type = factor(model_type, levels = c("Baseline","MS","MN","M1","M2","M3","MF")),
         task_type  = factor(task_type, c("Classification","Recommendation","Regression")),
         across(all_of(trd_cols), as.integer))

table(d$model_type, d$n_trd)     # exposes the structural confound
```

The cross-tabulation shows **aliasing**: baselines always use 0 training
sets, `MS` always 1, `Mx` span 2--8. So "number of datasets",
"single-vs-multi" and "baseline-vs-transfer" cannot be separated
globally. We isolate each effect on the appropriate subset.

# Exploratory data analysis

```{r eda, fig.width=12, fig.height=3.2}
het <- d |> group_by(TeD, task_type) |>
  summarise(mean=mean(score), sd=sd(score), min=min(score), max=max(score),
            n=n(), .groups="drop"); het
cat(sprintf("within-TeD SD ranges %.3f-%.3f  (%.1fx)\n", min(het$sd), max(het$sd),
            max(het$sd)/min(het$sd)))

p1 <- ggplot(d, aes(score)) + geom_histogram(bins=40, fill="#34495e", colour="white",
      linewidth=.1) + labs(x="metric score", y="count", title="(a) Response bounded in [0,1]")
p2 <- ggplot(d, aes(TeD, score, fill=task_type)) + geom_boxplot(outlier.size=.3, linewidth=.3) +
  scale_fill_manual(values=c(Classification="#2471a3",Recommendation="#e67e22",Regression="#27ae60")) +
  labs(x=NULL, y="metric score", fill=NULL, title="(b) Scale differs ~15x across test sets")

(p1 | p2)
```

# Large preliminary model and ANOVA

We fit a large model with all available factors and key interactions and
rank effects with **Type-II** sums of squares (order-independent under
the unbalanced design).

```{r anova-first}
m_big <- lm(score ~ TeD * model_type + TeD:n_trd, data = d)
Anova(m_big, type = 2) |> tidy() |>
  mutate(pct_SS = round(100*sumsq/sum(sumsq), 1)) |> arrange(desc(pct_SS))
```

`TeD` alone is \~93% of the sum of squares. Two consequences drive the
rest of the analysis: (1) a high overall `R^2` is uninformative (it just
reflects task differences), so we must **control for `TeD`**, and, (2)
because the test sets differ in scale by \~15x, raw effects are **not
comparable across `TeD`**.

# Transform for homogeneity: within-`TeD` standardization

Standardizing the score within each test set removes the between-`TeD`
scale and ensures **variance homogeneity** (every `TeD` has SD 1), so an
effect can be read as "how many SD of *that* test set". This is our
primary response. A clamped **logit**, which respects the `[0,1]` bound,
is used later as a robustness check.

```{r primary-model}
m_raw <- lm(score ~ TeD * model_type, data = d)   # natural scale (for intuition)
m_z   <- lm(z     ~ TeD * model_type, data = d)   # PRIMARY: SD-of-that-TeD units
summary(m_z)$adj.r.squared # expected to be lower than the raw R^2 because z already removed the giant TeD variance, leaving the interesting part to be explained
```

## Goodness of fit

```{r diagnostics, fig.height=3}
par(mfrow = c(1, 3))
plot(m_raw, which = 1, main = "(a) raw: variance funnel") # funnel -> spread grows with the fitted value = heteroscedasticity
# (homoscedasticity = residual variance constant across fitted values, violated here)
plot(m_z,   which = 1, main = "(b) per-TeD std: even") # spread is now expected to be even across fitted values (assumption satisfied)
plot(m_z,   which = 2, main = "(c) std model Q-Q") # Normal Q-Q: residuals sorted vs what a Normal predicts; on the diagonal = ~Normal
par(mfrow = c(1, 1))
```

The variance funnel of the raw model flattens after per-`TeD`
standardization. The Q-Q is straight apart from a mild left tail from
some genuine near-zero scores in `TeD7`. This is not a concern at
`n = 2996`, where the Central Limit Theorem makes coefficient inference
robust to non-Normality, and any residual non-constant variance does not
change our conclusions, which we confirm with the logit and
quasi-binomial models.

## Robustness models

We refit the same model two other ways, respecting the [0,1] bound, to
later check that the conclusions don't depend on the standardization
trick.

```{r robustness}
m_logit <- lm(logit ~ TeD * model_type, data = d) # models the logit of the score, unsquishing the scale, with an ordinary linear model
m_qb    <- glm(pmin(pmax(score,eps),1-eps) ~ TeD * model_type,
               family = quasibinomial(), data = d) # a generalized linear model, keeps the predictions inside [0,1] and assumes proportion-like variance with a flexible dispersion
cat("quasi-binomial dispersion:", round(summary(m_qb)$dispersion,3), "\n") # the dispersion (how spread-out the data is vs. a strict binomial)
```

# RQ1 - Does transfer learning improve over typical non-transfer learning?

## Transfer - Baseline comparison

Effect sizes are differences in **SD units** of each test set, with 95%
confidence intervals.

```{r rq1}
emm <- emmeans(m_z, ~ model_type | TeD) # adjusted mean (in SD-of-TeD units) for each architecture, within each test set

contrast(emm, method = list("TR - BL" = c(-1, rep(1/6, 6))), by = "TeD") |>
  summary(infer = TRUE) # weights are aligned to model_type: Baseline, MS, MN, M1, M2, M3, MF -> -1 on Baseline, 1/6 on each of the 6 transfer-based ones -> avg transfer - baseline.

contrast(emmeans(m_z, ~ model_type), method = list("TR - BL" = c(-1, rep(1/6, 6)))) |>
  summary(infer = TRUE) # same contrast, but on EMMs pooled over all test sets (single overall SD figure)

contrast(emmeans(m_z, ~ model_type), method = "trt.vs.ctrl", ref = 1) |>
  summary(infer = TRUE) # Dunnett ("Treatment vs Control") -> each architecture - baseline control, pooled over test sets

m_raw_grp <- lm(score ~ TeD * model_group, data = d) # separate raw score model using the BL/TR grouping to see real-scale numbers
emmeans(m_raw_grp, ~ model_group | TeD) |> summary() # baseline vs transfer mean score per test set
```

Transfer beats baseline overall (\~1.88 SD) and on **six of the seven**
test sets. Only `TeD4` shows no benefit. The Dunnett contrasts order the
architectures `MS < MN < M1 < M2 < M3 < MF`, all above baseline, meaning
that more refinement leads to larger gain.

We observe that the SE for the baseline is 0.041, whereas the transfer's
is 0.0034 (12x more precise). This is because the baseline rests on only
3 runs per test set, while the transfer has hundreds. This is also the
reason for the wider confidence intervals.

```{r rq1-fig, fig.width=7, fig.height=5}
# (a) raw EMMs: baseline vs transfer per test set
raw_emm <- as.data.frame(emmeans(m_raw_grp, ~ model_group | TeD))
pA <- ggplot(raw_emm, aes(TeD, emmean, colour = model_group)) +
  geom_point(position = position_dodge(.4), size = 2) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                position = position_dodge(.4), width = .25) +
  scale_colour_manual(values = c(BL = "#c0392b", TR = "#2471a3")) +
  labs(x = NULL, y = "mean score (raw)", colour = NULL,
       title = "(a) Raw scores: some gaps look negligible")

# (b) transfer - baseline per test set, in SD units
rq1_sd <- as.data.frame(summary(
  contrast(emm, method = list("TR - BL" = c(-1, rep(1/6, 6))), by = "TeD"),
  infer = TRUE)) |>
  mutate(signif = lower.CL > 0 | upper.CL < 0)
pB <- ggplot(rq1_sd, aes(TeD, estimate, fill = signif)) +
  geom_col(alpha = .85) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = .25) +
  geom_hline(yintercept = 0, linewidth = .3) +
  scale_fill_manual(values = c(`TRUE` = "#27ae60", `FALSE` = "#7f8c8d"), guide = "none") +
  labs(x = NULL, y = "transfer - baseline (SD of that TeD)",
       title = "(b) In context: transfer helps on 6/7")

(pA / pB)
```

## Transfer-model refinement depth analysis

```{r rq1-depth}
md <- d |> filter(model_type %in% c("MN","M1","M2","M3","MF"))
emtrends(lm(z ~ TeD + depth, data = md), ~ 1, var = "depth") |> summary(infer = TRUE)
```

Filtering to the multi-source models, where `depth` runs 0-4, and
fitting a linear model that controls for `TeD`, we read off the one
common slope = SD gained per refined part (0.28, CI [0.26, 0.31]).

```{r rq1-figb, fig.width=7, fig.height=3}
# (b) each architecture vs baseline (Dunnett), SD units
dunn <- as.data.frame(summary(
  contrast(emmeans(m_z, ~ model_type), method = "trt.vs.ctrl", ref = 1), infer = TRUE)) |>
  mutate(arch = factor(sub(" - Baseline", "", contrast),
                       levels = c("MS","MN","M1","M2","M3","MF")))
pD <- ggplot(dunn, aes(arch, estimate)) +
  geom_col(fill = "#2471a3", alpha = .85) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = .25) +
  labs(x = NULL, y = "gain vs baseline (SD units)",
       title = "Deeper refinement, larger gain")

pD
```

# RQ2 - How is the model performance affected by the choice of training datasets?

## Number of training datasets (multi-`TrD` subset)

We only consider the 5 multi-source models where `n_trd` ranges from 2-8
freely, breaking the baseline=0 and MS=1 confound. `n_c` represents
"datasets above/below average", making the intercept and slope
interpretable. We also fit a richer model with a square term I(n_c\^2),
allowing the line to bend, then anova (small, big) tests "does the curve
fit significantly better than the straight line?" or technically "does
adding I(n_c\^2) drop the residual error by more than chance?".
Significance, in this case, means that the straight line is too simple
and a curvature is real. To know that it is diminishing, we check the
sign of the I(n_c\^2) coefficient: negative = the curve bends downward =
the slope shrinks as n_c grows = each extra dataset helps less =
diminishing returns.

```{r rq2-number}
mx <- d |> filter(model_type %in% c("M1","M2","M3","MF","MN")) |>
  mutate(n_c = n_trd - mean(n_trd))

m_num <- lm(z ~ TeD + n_c, data = mx) # linear model, using + (one common slope, no interaction), TeD is here to control the task, n_c carries the effect of interest

emtrends(m_num, ~ 1, var = "n_c") |> summary(infer = TRUE) # "how much z changes per dataset" -> overall slope (different heights for each test set) because there is no TeD:n_c interaction

anova(m_num, lm(z ~ TeD + n_c + I(n_c^2), data = mx))

m_quad <- lm(z ~ TeD + n_c + I(n_c^2), data = mx)
summary(m_quad)        # looking at the I(n_c^2) row, "Estimate" column
```

We notice that adding the term drops the residual error (RSS) from
1821.7 to 1804.4, a reduction of 17.3 for 1 extra parameter, giving
F=25.3 at p = 5.3e-07. So, the curve term is significant, although a
minor part of the variance (17 out of 1822), which is what "diminishing
returns" are, a subtle bend. The I(n_c\^2) coefficient is -0.025,
therefore the curve is concave and the returns are diminishing. The
linear term is 0.14 (SD per dataset on average).

```{r rq2-fig, fig.width=7, fig.height=3}
# (a) number-of-datasets trend (SD units, averaged over TeD)
obs <- mx |> group_by(TeD, n_trd) |> summarise(z = mean(z), .groups = "drop") |>
  group_by(n_trd) |> summarise(z = mean(z), .groups = "drop")
newd <- expand.grid(TeD = levels(mx$TeD), n_trd = sort(unique(mx$n_trd)))
newd$n_c  <- newd$n_trd - mean(mx$n_trd)
newd$pred <- predict(m_quad, newdata = newd)
fit <- newd |> group_by(n_trd) |> summarise(pred = mean(pred), .groups = "drop")
pC <- ggplot() +
  geom_point(data = obs, aes(n_trd, z), size = 2, colour = "#34495e") +
  geom_line(data = fit, aes(n_trd, pred), colour = "#2471a3", linewidth = 1) +
  labs(x = "number of training datasets", y = "score (SD units, avg over TeD)",
       title = "More datasets help, diminishing")

pC
```

## Which training datasets help or hurt and when

We filter out the non-transfer runs to keep the subset where the eight
`TrD` vary. Next, we fit a linear model to a formula that gives each
dataset a main effect and a `TeD` interaction and observe the effects
(slopes). ANOVA also tells us if a more complex model that allows
datasets to have a distinct effect depending on the target provides a
better fit for the data or not, therefore a dataset's usefulness depends
on the test set.

```{r rq2-which}
tr   <- d |> filter(n_trd >= 1)
m_id <- lm(as.formula(paste("z ~ TeD * (", paste(trd_cols, collapse=" + "), ")")), data = tr) # z ~ TeD * (TrD1 + ... + TrD8)
incl <- map_dfr(trd_cols, function(k){ # loop the function over each dataset name k
  et <- summary(emtrends(m_id, ~ 1, var = k), infer = TRUE) # the slope of z w.r.t. TrD k, i.e. the effect of including/excluding the dataset, since TrD is coded 1/0
  tibble(TrD=k, delta_SD=et[[paste0(k,".trend")]], lower=et[["lower.CL"]],
         upper=et[["upper.CL"]], p=et[["p.value"]]) # one row result
}) |> arrange(desc(delta_SD)); incl # most helpful first

anova(lm(as.formula(paste("z ~ TeD +", paste(trd_cols, collapse="+"))), data=tr), m_id) # ANOVA to compare the 2 and see whether a dataset's usefulness depends on the target task -> it does

# per-test-set inclusion effect
for (k in trd_cols) print(summary(emtrends(m_id, ~ TeD, var = k), infer = TRUE))
```

We observe that including `TrD8/TrD5/TrD7` helps most.
`TrD2/TrD3/TrD6/TrD4` help mildly. **`TrD1` hurts**. The `TrD - TeD`
interaction is highly significant, cutting the RSS from 2028.7 to 1616.0
(a difference of 412.7 points or 20% of Model 1's RSS) for those 48
terms, with a significant F = 15.5 at p \< 2.2e-16: dataset usefulness
depends on the target task.

```{r rq2-which-fig, fig.width=4, fig.height=2.8}
ggplot(incl, aes(delta_SD, reorder(TrD, delta_SD), colour = delta_SD > 0)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = .2) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(`TRUE` = "#27ae60", `FALSE` = "#c0392b"), guide = "none") +
  labs(x = "effect of including dataset (SD units, avg over TeD)", y = NULL,
       title = "Which training sets help vs. hurt")
```

# Robustness: do conclusions hold on the bounded (logit) scale?

We define a reusable function that computest the per-test-set transfer -
baseline contrast, with the same weights as in RQ1. Then, we match the
m_z, m_logit and m_qb tables on TeD, renaming columns appropriately, and
then flag whether all three agree in the direction per test set, i.e.
the robustness verdict.

```{r robust-check}
chk <- function(model) contrast(emmeans(model, ~ model_type | TeD),
        method = list(d = c(-1, rep(1/6,6))), by="TeD") |> summary() |>
        transmute(TeD, est = estimate)
chk(m_z)    |> rename(SD_units = est) |>
  left_join(chk(m_logit) |> rename(logit    = est), by = "TeD") |>
  left_join(chk(m_qb)    |> rename(quasibin = est), by = "TeD") |>
  mutate(sign_agrees = sign(SD_units) == sign(logit) &
           sign(SD_units) == sign(quasibin))
```

The results show that the modelling did not distort the results. 6 of 7
test sets are `TRUE`, showing that on each one there is a real effect,
agreed by 3 different ways of computation. `TeD4` = `FALSE` is a
non-issue because, looking at its -0.003, +0.004, -0.003 values, they
are essentially zero, flipping harmlessly between methods, and we can
easily mark this as noise. This disagreement actually reinforces our
RQ1, with `TeD4` having no real effect.

# Session info

We include the R version and every loaded package and version to ensure
reproducibility.

```{r session}
sessionInfo()
```
