---
title: "DSAIT4200 Coursework B --- Analysis of a Transfer-Learning Experiment"
subtitle: "Reproducible analysis (source for the report)"
author: "Alex"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 3, dpi = 130)
```

# Purpose and approach

The experiment measures a **metric score in `[0, 1]`** for nine *models* on seven
*test datasets* (`TeD`); transfer-learning models are trained on subsets of eight
*training datasets* (`TrD1..TrD8`). We answer:

* **RQ1** -- does transfer learning beat non-transfer baselines?
* **RQ2** -- how do the *number* and *identity* of training datasets, and the
  *depth of refinement*, affect performance?

Following the *Frequentist Statistical Testing and Modeling* lectures and the
hands-on session, we (1) explore the data; (2) fit a **large preliminary model
and rank effects with Type-II ANOVA** to guide the analysis; (3) **transform the
response per test dataset** (within-`TeD` standardization) so effects are
**comparable across test sets** -- a raw difference that is negligible on a
high-variance test set can be large *in the context of* a low-variance one;
(4) build structure-aware models and report **effect sizes with confidence
intervals** (in SD units), not just p-values; and (5) confirm everything with a
bounded-response **logit / quasi-binomial** model.

```{r libs}
## install.packages(c("tidyverse","car","emmeans","broom"))
library(tidyverse); library(car); library(emmeans); library(broom)
theme_set(theme_bw(base_size = 11)); emm_options(rg.limit = 200000)
```

# Data preparation

We derive the *non-explicit* factors the design hides: the **number** of training
datasets `n_trd`; the architecture `model_type` (Baseline as reference);
`model_group` (Baseline vs Transfer); the broad `task_type`; the **refinement
depth** of the multi-`TrD` models (`MN`=0 ... `MF`=4); the **within-`TeD`
standardized** score `z` (primary response, SD-of-that-TeD units); and a clamped
**logit** for the bounded-response robustness check.

```{r prep}
d <- readr::read_csv("data.csv", show_col_types = FALSE)
trd_cols <- paste0("TrD", 1:8); eps <- 0.005
d <- d |>
  mutate(
    n_trd       = rowSums(across(all_of(trd_cols))),
    model_group = if_else(model %in% c("B1","B2","B3"), "BL", "TR"),
    model_type  = if_else(model %in% c("B1","B2","B3"), "Baseline", model),
    task_type   = case_when(TeD %in% paste0("TeD",1:4) ~ "Classification",
                            TeD == "TeD5" ~ "Recommendation", TRUE ~ "Regression"),
    depth       = case_when(model_type=="MN"~0, model_type=="M1"~1, model_type=="M2"~2,
                            model_type=="M3"~3, model_type=="MF"~4, TRUE~NA_real_),
    logit       = qlogis(pmin(pmax(score, eps), 1 - eps))) |>
  group_by(TeD) |> mutate(z = (score - mean(score)) / sd(score)) |> ungroup() |>
  mutate(TeD        = factor(TeD, levels = paste0("TeD",1:7)),
         model_group= factor(model_group, levels = c("BL","TR")),
         model_type = factor(model_type, levels = c("Baseline","MS","MN","M1","M2","M3","MF")),
         task_type  = factor(task_type, c("Classification","Recommendation","Regression")),
         across(all_of(trd_cols), as.integer))
table(d$model_type, d$n_trd)     # exposes the structural confound (see report)
```

The cross-tabulation shows **aliasing**: baselines always use 0 training sets,
`MS` always 1, `Mx` span 2--8. So "number of datasets", "single-vs-multi" and
"baseline-vs-transfer" cannot be separated globally; we isolate each effect on
the appropriate subset.

# Exploratory data analysis

```{r eda}
het <- d |> group_by(TeD, task_type) |>
  summarise(mean=mean(score), sd=sd(score), min=min(score), max=max(score),
            n=n(), .groups="drop"); het
cat(sprintf("within-TeD SD ranges %.3f-%.3f  (%.1fx)\n", min(het$sd), max(het$sd),
            max(het$sd)/min(het$sd)))

p1 <- ggplot(d, aes(score)) + geom_histogram(bins=40, fill="#34495e", colour="white",
      linewidth=.1) + labs(x="metric score", y="count", title="(a) Response bounded in [0,1]")
p2 <- ggplot(d, aes(TeD, score, fill=task_type)) + geom_boxplot(outlier.size=.3, linewidth=.3) +
  scale_fill_manual(values=c(Classification="#2471a3",Recommendation="#e67e22",Regression="#27ae60")) +
  labs(x=NULL, y="metric score", fill=NULL, title="(b) Scale differs ~15x across test sets")
print(p1); print(p2)
```

# Large preliminary model and ANOVA (guides the analysis)

We fit a large model with all available factors and key interactions and rank
effects with **Type-II** sums of squares (order-independent under the unbalanced
design).

```{r anova-first}
m_big <- lm(score ~ TeD * model_type + TeD:n_trd, data = d)
Anova(m_big, type = 2) |> tidy() |>
  mutate(pct_SS = round(100*sumsq/sum(sumsq), 1)) |> arrange(desc(pct_SS))
```

`TeD` alone is ~93% of the sum of squares. Two consequences drive the rest of the
analysis: (i) a high overall `R^2` is *uninformative* (it just reflects task
differences), so we must **control for `TeD`**; and (ii) because the test sets
differ in scale by ~15x, raw effects are **not comparable across `TeD`** -- which
motivates the transform below.

# Transform for homogeneity: within-`TeD` standardization

Standardizing the score within each test set removes the between-`TeD` scale and
yields **perfect variance homogeneity** (every `TeD` has SD 1), so an effect can
be read as "how many standard deviations of *that* test set". This is our primary
response. (A clamped **logit**, which respects the `[0,1]` bound, is used later as
a robustness check.)

```{r primary-model}
m_raw <- lm(score ~ TeD * model_type, data = d)   # natural scale (for intuition)
m_z   <- lm(z     ~ TeD * model_type, data = d)   # PRIMARY: SD-of-that-TeD units
summary(m_z)$adj.r.squared
```

## Goodness of fit

```{r diagnostics, fig.height=3}
par(mfrow = c(1, 3))
plot(m_raw, which = 1, main = "(a) raw: variance funnel")
plot(m_z,   which = 1, main = "(b) per-TeD std: even")
plot(m_z,   which = 2, main = "(c) std model Q-Q")
par(mfrow = c(1, 1))
car::ncvTest(m_raw); car::ncvTest(m_z)   # heteroscedasticity: raw severe, std much milder
```

The variance funnel of the raw model disappears after per-`TeD` standardization;
the Q--Q is straight apart from a mild left tail from ~4 genuine near-zero scores
in `TeD7`. With `n = 2996`, coefficient inference is robust by the CLT.

## Robustness models

```{r robustness}
m_logit <- lm(logit ~ TeD * model_type, data = d)                       # respects [0,1]
m_qb    <- glm(pmin(pmax(score,eps),1-eps) ~ TeD * model_type,
               family = quasibinomial(), data = d)                      # proportion GLM
cat("quasi-binomial dispersion:", round(summary(m_qb)$dispersion,3), "\n")
```

# RQ1 --- Does transfer learning beat the baselines?

Effect sizes are differences in **SD units** of each test set, with 95% CIs.

```{r rq1}
emm <- emmeans(m_z, ~ model_type | TeD)
## Transfer (equal-weighted avg of the 6 architectures) - Baseline, per test set
contrast(emm, method = list("TR - BL" = c(-1, rep(1/6, 6))), by = "TeD") |>
  summary(infer = TRUE)
## overall, averaged over test sets
contrast(emmeans(m_z, ~ model_type), method = list("TR - BL" = c(-1, rep(1/6, 6)))) |>
  summary(infer = TRUE)
## Dunnett: every architecture vs the Baseline control (avg over TeD)
contrast(emmeans(m_z, ~ model_type), method = "trt.vs.ctrl", ref = 1) |>
  summary(infer = TRUE)
## raw-score EMMs for intuitive magnitudes (reported alongside in the paper)
emmeans(m_raw, ~ model_group | TeD) |> summary()
```

Transfer beats baseline overall (~1.9 SD) and on **six of seven** test sets;
only `TeD4` shows no benefit. The Dunnett contrasts order the architectures
`MS < MN < M1 < M2 < M3 < MF`, all above baseline -- more refinement, larger
gain. Baseline EMMs have wide CIs (only 3 points per test set): a key caveat.

# RQ2 --- Effect of the choice of training datasets

## (a) Number of training datasets (multi-`TrD` subset)

```{r rq2-number}
mx <- d |> filter(model_type %in% c("M1","M2","M3","MF","MN")) |>
  mutate(n_c = n_trd - mean(n_trd))
m_num <- lm(z ~ TeD + n_c, data = mx)
emtrends(m_num, ~ 1, var = "n_c") |> summary(infer = TRUE)            # SD per added dataset
anova(m_num, lm(z ~ TeD + n_c + I(n_c^2), data = mx))                 # diminishing returns?
```

## (b) Which training datasets help or hurt, and TrD x TeD interaction

```{r rq2-which}
tr   <- d |> filter(n_trd >= 1)
m_id <- lm(as.formula(paste("z ~ TeD * (", paste(trd_cols, collapse=" + "), ")")), data = tr)
incl <- map_dfr(trd_cols, function(k){
  et <- summary(emtrends(m_id, ~ 1, var = k), infer = TRUE)
  tibble(TrD=k, delta_SD=et[[paste0(k,".trend")]], lower=et[["lower.CL"]],
         upper=et[["upper.CL"]], p=et[["p.value"]])
}) |> arrange(desc(delta_SD)); incl
## does each dataset's effect depend on the test set?
anova(lm(as.formula(paste("z ~ TeD +", paste(trd_cols, collapse="+"))), data=tr), m_id)
```

Including `TrD8/TrD5/TrD7` helps most; `TrD2/TrD3/TrD6/TrD4` help mildly;
**`TrD1` hurts**. The `TrD x TeD` interaction is highly significant: dataset
usefulness depends on the target task.

## (c) Refinement depth

```{r rq2-depth}
md <- d |> filter(model_type %in% c("MN","M1","M2","M3","MF"))
emtrends(lm(z ~ TeD + depth, data = md), ~ 1, var = "depth") |> summary(infer = TRUE)
```

# Robustness: do conclusions hold on the bounded (logit) scale?

```{r robust-check}
chk <- function(model) contrast(emmeans(model, ~ model_type | TeD),
        method = list(d = c(-1, rep(1/6,6))), by="TeD") |> summary() |>
        transmute(TeD, est = estimate)
left_join(chk(m_z) |> rename(SD_units=est), chk(m_logit) |> rename(logit=est), by="TeD") |>
  mutate(sign_agrees = sign(SD_units) == sign(logit))
```

# Session info

```{r session}
sessionInfo()
```
